The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)

### 2025-02-15 paper.tex
Theorem provers, formal model

\bibliographystyle{plainurl}
\title{The Choose-Your-Own-Adventure Calculus} % (Pearl)

@@

Some of the most remarkable results in mathematics reveal surprising connections between different
branches of the discipline. The aim of this paper is to point out a modest, but still
remarkable, similarity between a range of different interactive programming systems.

@@

\item The calculus allows us to make the aforementioned subtle differences precise. We define the
notions of \emph{correctness} and \emph{completeness} for the choose-your-own-adventure calculus.
To distinguish the different ways of embedding the interactions in the edited programs, we
also formally define \emph{base-level} and \emph{meta-level} integration of the system.

@@

@@

@@

@@

@@

>>>>

\newcommand{\ident}[1]{\textsf{#1}}
\newcommand{\sigmaN}{\sigma_0}
\newcommand{\select}{\textnormal{\ident{choose}}}
\newcommand{\preview}{\textnormal{\ident{preview}}}
\newcommand{\links}{\textnormal{\ident{links}}}
\newcommand{\choices}{\textnormal{\ident{choices}}}
\newcommand{\op}{\textit{op}}
\newcommand{\argmax}{\textit{arg}\,\textit{max}\,}

\bibliographystyle{plainurl}
\title{The Choose-Your-Own-Adventure Calculus (Pearl/Brave New Idea)}

@@

Some of the most remarkable results in mathematics reveal connections between different
branches of the discipline. The aim of this paper is to point out a modest, but still
remarkable, similarity between a range of different interactive programming systems.

@@

\item The calculus allows us to make the aforementioned subtle differences precise. We define the
notions of \emph{correctness} and \emph{completeness} for the choose-your-own-adventure calculus.
To distinguish the different ways of embedding the interactions in the edited programs, we
also formally define \emph{internal} and \emph{external} mode of system integration.

@@

\begin{figure}[t]
  \includegraphics[width=0.7\textwidth]{fig/worldbank.png}
  \caption{F\# code editor showing completions offered by the World Bank type provider.}
  \label{fig:worldbank}
\end{figure}

@@

Computer scientists studying programming have long focused on programming languages as syntactic
entities, sometimes neglecting the interactive environments in which they are inevitably
embedded \cite{rpg-2012-revolution}. Notably, in many of the motivating examples that we draw
from in this section, the interactive aspect of the system is only described in supplementary
materials \cite{brady-2015-idris,syme-2013-inforich,altenkirch-1994-alf}. Only recently, programming
language theory started to be used to study interactive environments
\cite{adams-2025-grove,mayer-2018-bidirectional}. Our work contributes to this research direction.

The following sections review four different instances of the choose-your-own-adventure
interaction pattern. In all of those, an interactive editor offers the user some kind of~a~com\-pletion list during working with the system.

\subparagraph{Type providers}

F\# type providers \cite{syme-2013-inforich} are a mechanism for integrating external data
sources into the F\# type system. A type provider is a compiler extension, loaded and
executed at compile-time and at edit-time. It can run arbitrary code to read the structure of
external data and use it to generate a suitable statically-typed representation of the
data, typically as objects with members. Type providers can, for example, infer the type from a
sample JSON \cite{petricek-2016-fsdata} or read a database schema.

The example in Figure~\ref{fig:worldbank} shows a simple type provider for accessing information
from the World Development Indicators database. The provided \ident{wb} object allows the programmer
to access any indicator of any country in the database by choosing an appropriate \ident{[Country]}
and an \ident{[Indicator]} in a chain of members
\ident{wb}.\ident{Countries}.\ident{[Country]}.\ident{Indicator}.\ident{[Indicator]}.
The result is a time series with values for the given indicator and a country. More generally,
the example can be seen as a special case of a type provider for slicing n-dimensional
data cube \cite{petricek-2022-thegamma} -- we choose a fixed value for two of the three
dimensions (country, indicator, time).

When using the type provider, the user types the first line of code and triggers auto-completion
by typing \ident{wb} followed by the dot. The rest of the code is constructed by choosing an
option from a list and typing another dot.\footnote{This interaction pattern has been
lightheartedly called \emph{dot-driven development} by Phil Trelford \cite{seemann-2021-head}.}

\begin{figure}[t]
\includegraphics[width=0.48\textwidth]{fig/thegamma1.png}~~
\includegraphics[width=0.48\textwidth]{fig/thegamma2.png}
\caption{Constructing a query in The Gamma. We count the number of gold and silver medals for
each athlete and sort the data by the number of gold medals.}
\label{fig:thegamma}
\end{figure}

\subparagraph{Data exploration}

The Gamma~\cite{petricek-2022-thegamma} is a programmatic data exploration environment
for non-programmers. In The Gamma, type providers are the primary programming mechanism. They
are used not just for data access, but also for constructing queries.

The type provider shown in Figure~\ref{fig:thegamma} lets the user construct an SQL-like query by
repeatedly choosing operations and their parameters \cite{petricek-2017-dotdriven}. It keeps track
of the schema and uses it to generate all possible valid parameters. When sortin data, it generates
an object with two members for each columns -- one for ascending and one for descending sort.
Similarly, the grouping operation first offers all columns as possible grouping keys and then
lets the user choose from a range of pre-defined aggregations (sum, count, average, concatenate).
The system also evaluates the query on the fly, providing a live preview during editing \cite{petricek-2020-live}.

The interaction pattern is the same as before. After the usertriggers auto-completion, they
repeatedly select an operation and its parameters to construct a query. One notable difference
is that the structure of the generated types is potentially infinite (the user can keep adding
further operations) and so the types are generated lazily.

\subparagraph{AI assistants}

The third instance of the choose-your-own-adventure interaction pattern comes from the work on
semi-automatic data wrangling tools known as AI assistants \cite{petricek-2023-aias}.
An AI assistant guides the analyst through a data wrangling problem such as reconciling mismatched
datasets, filling missing values or inferring data format and types. An AI assistant solves
the problem automatically and suggests an initial data transformation, but it also generates a
number of constraints that the user can choose from to refine the initial solution. If the initial
solution is not correct, the user chooses a constraint and the AI assistant runs again, suggesting
a new data transformation that respects the constraint.

Figure~\ref{fig:aia} shows an example. It uses the datadiff \cite{sutton-2018-datadiff} AI
assistant, running in a Wrattler notebook \cite{petricek-2018-wrattler}, to merge broadband
quality data published by Ofcom for two subsequent years. The format of the CSV files for the
two years differs. Columns were added, removed, renamed and their order has changed. In the
example, we selected 6 columns from the year 2015 and want to find matching data from 2014.

When the AI assistants runs automatically, it correctly maps the numerical columns, but it
incorrectly maps the \ident{Urban.rural} (2014) column to \ident{Nation} (2015). This happens
because both columns are categorical and have three values with similar distribution. A data
analyst can easily spot the mistake. They click the ``+'' button to add a constraint and choose
\ident{Don't match Urban.rural and Nation} to specify that the two columns should not be matched.
Datadiff then runs again and finds the correct matching.

\begin{figure}[t]
  \includegraphics[width=0.9\textwidth]{fig/aia.png}
  \caption{Using the datadiff AI assistant to reconcile the structure of the two datasets.
    The user is offered a list of constraints to prevent or force matching between specific columns.}
  \label{fig:aia}
\end{figure}

@@

The interaction patter is the same as in the previous two cases. The analyst constructs the
correct data transformation by repeatedly choosing from a list of options, until they obtain
the desired result. However, the way the interaction pattern is implemented differs.
First, in the case of type providers, we are gradually constructing a program by adding operations to
a method chain. Now, the AI assistant synthesizes a data transformation (program) and we are
gradually adding constraints to control the synthesis. Second, in the case of type providers,
the completion list offered all possible members of the object. Now, the list offers
constraints recommended by the AI assistant which may not be complete.

\begin{figure}[t]
  \includegraphics[width=0.6\textwidth]{fig/alf.png}
  \caption{Constructing a proof of the transitivity of the $\leq$ relation in the \textsc{Alf}
    editor. The user is offered a range of variables and constructors in scope at the current
    location. \cite{altenkirch-1994-alf}}
  \label{fig:alf}
\end{figure}

\subparagraph{Interactive theorem provers}
A fourth example of the choose-your-own-adventure interactive pattern can be found in interactive
theorem provers. When writing programs in systems like Idris \cite{brady-2021-idris2}, the
user typically works by stating the desired conclusion and filling the implementation with a hole.
The system provides a range of interactive editing capabilities to fill the holes \cite{mcbride-1999-dependently}.
It can, for example, generate a case split or search for a proof \cite{brady-2015-idris}.

Systems like Idris provide key bindings to invoke the completions, but the functionality could
also be offered through a user interface. An example that illustrates this is the interactive editor
for the \textsc{Alf} theorem prover \cite{magnusson-1994-alf}, which is based on the refinement of
an incomplete proof object \cite{altenkirch-1994-alf}. This is illustrated in Figure~\ref{fig:alf}.
The user is proving the transitivity of the $\leq$ relation for Peano arithmetic natural numbers.
They pattern match on the proof argument $p$ and complete the first branch. For the second branch,
they need to fill a hole $?_{p2}$ (called a wildcard in \textsc{Alf}). They trigger a completion
and a pop-up menu shows the available variables and constructors, including \ident{leq\_trans} that
can be used to complete the proof. After choosing \ident{leq\_trans}, two new holes are generated
for its arguments. Those can be, again, filled interactively, by choosing $p_1$ and $p$ from the
completion.

The interaction pattern is again the same. The user repeatedly triggers a completion and uses
it to refine and complete their proof by filling holes. There are subtle differences too. Unlike
with AI assistants, each completion directly refines the proof that the user is editing. Unlike
with type providers, a completion may generate multiple new holes, rather than just adding to a
chain of operations.

The \textsc{Alf} editor is a historical example, but a similar user interface could be built for
systems like Idris or Coq. The two would work differently. As in \textsc{Alf}, Idris source code
represents the proof itself and a completion would replace a hole with a suggested term. In Coq,
the proof is a series of tactic invocations and so selected completions would be added to this
list and would form a trace of the interaction with the user.

@@

A system that implements the choose-your-own-adventure interaction pattern repeatedly offers
the user a range of options to choose from. Each of the options is designated by an identifier.
The system also maintains a state during the process which determines subsequent options. The state
may not be visible to the user, but the user can always explicitly request the program constructed
so far.

We can think of the interaction with the system as navigating through a tree structure, starting
from a root and choosing one of the possible branches in each step.\footnote{Serving as another
evidence for the surprising effectiveness of the concept of a tree \cite{nesetril-2005-strom}.}
In the following definition, the key $\choices$ operation can thus be seen as returning
branches of a given node.

\begin{definition}[Choose-your-own-adventure system]\label{def:calculus}
Given expressions $e\in \mathbb{E}$ and states $\sigma \in \Sigma$, a choose-your-own-adventure
system is a pair of operations \choices, \select\ such that:

\vspace{-0.25em}
\begin{itemize}
  \item $\choices(\sigma) = (\iota_1\mapsto\sigma_1, \ldots, \iota_n\mapsto\sigma_n)$ is
    an operation that takes a state and generates options designated by an identifier $\iota_i$
    and represented by a state $\sigma_i$,
  \item $\select(\sigma) &=& e$ is an operation that returns generated program for a given state.
\end{itemize}
\end{definition}

The definition is not a programming language calculus in the usual sense in that it does not
define a concrete syntax with reduction rules. It is an abstract algebraic structure that captures
the structure of a system that supports the choose-your-own-adventure interaction pattern.
The definition is close to that of an AI assistant \cite{petricek-2023-aias}, which is written
using a language specific for the data wrangling domain (such as cleaning scripts or input and
output data) but is structurally similar. It is also worth noting that the definition may describe
not just trees, but also graphs with cycles -- a system can return to an already visited state.
This is not practically useful, but it does not pose a theoretical problem.

\subparagraph{External mode of embedding}
One of the subtle questions about the choose-your-own-adventure pattern raised in the introduction
concerns the different ways in which a trace of the interaction is embedded in the interactively
constructed program. In the \emph{external mode}, the interaction results in code that becomes a
part of the edited program, but it is not possible to reconstruct the steps used to generate the code.

The choose-your-own-adventure interaction pattern is typically used to complete a partial program.
To model this, we assume that the host language has a notion of a hole, written as~$?$
and that a user can select a part of program to invoke the completion on. We write $E[e]$ for a
completion context, akin to evaluation contexts in operational semantics.

We assume that, for a program containing a hole in a completion context $E[?]$, we can construct
an initial choose-your-own-adventure state using an operation
$\ident{init}(E[?])=\sigmaN$.

\begin{definition}\label{def:external}
An expression $E[?]$ is completed as $E[e]$ via external embedding of an interaction with
a choose-your-own-adventure system consisting of \choices\ and \select\ if:

\vspace{-0.25em}
\raggedright
\begin{enumerate}
\item $\ident{init}(E[?]) = \sigmaN$ obtains the initial state of a choose-your-own-interaction system,
\item $\sigma_n$ is a system state such that $\forall i\in{1\ldots n}.(\iota_i\mapsto\sigma_i)\in\choices{(\sigma_{n-1})}$,
  i.e.,~the user\\ makes a series of choices resulting in a final state of the system $\sigma_n$,
\item $E[e]$ where $e=\select(\sigma_n)$, i.e.,~the final program is constructed by replacing the hole
  in the completion context with the expression $e$ generated from $\sigma_n$.
\end{enumerate}
\end{definition}

As we will see when we revisit the earlier examples formally, the external mode of embedding is
used, for example, in the case of interactive theorem provers like \textsc{Alf} or Idris.
In those systems, the user triggers the completion on a proof (program) containing a hole.
They then fill the hole and, possibly iteratively, further holes in the generated proof. The final
expression is embedded in the source code, but it does not indicate what options, identified by
$\iota_1, \ldots, \iota_n$, were selected in the process.

\subparagraph{Internal mode of embedding}
In the \emph{internal mode}, a trace of the interaction with a choose-your-own-adventure system
is embedded directly in the constructed program. This is the case with type providers, where a
user chooses a sequence of object members to be accessed. The same would be the case in a completion
system for Coq that would offer tactics to apply, becuase the resulting proof would contain a
record of the selected tactics.

To talk about the internal mode formally, we again need the \ident{init} operation, but also an
operation \ident{decode} that extracts identifiers of invoked completions from an expression.
An internal embedding is the same as external embedding with an additional constraint:

\begin{definition}\label{def:internal}
An expression $E[?]$ is completed as $E[e]$ via internal embedding of an interaction with
a choose-your-own-adventure system consisting of \choices\ and \select\ if:

\vspace{-0.25em}
\raggedright
\begin{enumerate}
\item $E[?]$ is completed as $E[e]$ via external embedding using \ident{init}, \choices\ and \select\\
  through a series of choices designated by identifiers $\iota_1, \ldots, \iota_n$,
\item it also holds that $\ident{decode}(e)=(\iota_1, \ldots, \iota_n)$.
\end{enumerate}
\end{definition}

If a choose-your-own-adventure system is integrated in a programming language
through internal embedding, we can reconstruct the choices through which the user constructed
an expression $e$ in a completion context $E[e]$, assuming they used the interactive system rather
than entering the code directly. This also means that we can reconstruct the final state $\sigma_n$
of the system by starting from $\ident{init}(E[?])$ and following the choices specified by
$\iota_1, \ldots, \iota_n$.

@@

The internal mode can also be used when
working with theorem provers.

In Coq, where proofs are represented as series of tactics,

However, this would also be case in a
completion engine for the Coq proof system that would offer a choice of tactics to apply, because
Coq proofs are represented

\subsection{Data exploration}

\subsection{AI assistants}

\subsection{Interactive theorem proving}
